---
title: "Final Project Part 3 Report"
author: 
  - "Gadiel David Flores"
  - "Wendy Huang"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
---

# Contributions
- **Gadiel David Flores:** Description of contributions.
- **Wendy Huang:** Description of contributions.

# Introduction

```{r include-intro, echo=FALSE, eval=FALSE}
# Write the introduction here, ensuring it addresses:
# 1. The motivation and relevance of the project.
# 2. The research question of interest.
# 3. Connection of the research question to at least three peer-reviewed papers.
# 4. Why linear regression is suitable to answer the research question.
```


# Methods

## 1. **Data Preparation**
- **Tools Used**: Analysis used **R** with `tidyverse` for data handling and `car` for diagnostics.   
- **Steps Taken**:  
  - Defined **New Housing Price Index** as the response variable and identified predictors like absorption, construction, GDP, and CPI.  
  - Removed missing values and duplicates, merged data from 1997-2016, and divided it quarterly.  
  - Used Cook's Distance to remove influential outliers.  
These steps ensured a clean dataset for accurate regression analysis.

## 2. **Exploratory Data Analysis (EDA)**
- **Tools Used**: `ggplot2` for visualization and `corrplot` for visualizing correlations.  
Scatterplots were used to check for linearity between predictors and the response variable. Some predictors showed potential relationships and multicollinearity issues. Variance inflation factors (VIFs) flagged these predictors with high multicollinearity for further review. This exploratory data analysis provided essential insights into variable relationships, guiding decisions on transformations and feature selection for improved model accuracy.

## 3. **Model Building**
- **Tools Used**: `MASS` for stepwise regression and `caret` for cross-validation.  
Both AIC- and BIC-based stepwise regressions tested multiple models to identify the most efficient predictor combinations.Log transformations were used for skewed predictors like `Detached_Unabsorbed_Quarterly_Avg`, improving linearity and stabilizing variance. A Box-Cox transformation (\( \lambda = -0.3434 \)) was applied to the response variable, addressing residual non-normality and heteroscedasticity. These steps balanced model simplicity with predictive accuracy for optimal performance.

## 4. **Assessment of Assumptions**
- **Tools Used**: Residual plots, Q-Q plots, and statistical tests like Cook’s Distance.  
The model's assumptions were thoroughly checked: linearity and homoscedasticity were verified using residuals vs. fitted value and residual vs. predictors plots. Normality of residuals was assessed with Q-Q plots and histograms, revealing minor tail deviations. These steps ensured the regression model's validity, enhancing confidence in its results and inferences.

## 5. **Model Diagnostics**
- **Tools Used**: Performance metrics such as Adjusted \( R^2 \), AIC, BIC, RMSE, and MAE.  
Model performance was compared using AIC and BIC criteria. The simpler 8-predictor BIC model was chosen for its nearly equivalent performance to the more complex AIC model. These diagnostics ensured the selected model effectively balanced simplicity and performance, adhering to best practices in evaluating regression models.

## 6. **Mitigating Issues**
- **Tools Used**: Log transformations and feature selection techniques.  
To enhance model reliability, predictors with high VIFs, such as `GDP_Quarterly_Avg`, were excluded to reduce multicollinearity. Log transformations addressed non-linearity in variables like `Completed_Construction_Semi`. Outliers identified through Cook’s Distance were removed to minimize their impact on coefficients. These actions ensured the model adhered to regression assumptions and improved its interpretability and accuracy.

## 7. **Conclusion of Methods**
The stepwise, structured approach to developing the MLR model ensured robustness and validity while aligning with theoretical principles from the course material. Preprocessing and feature selection techniques addressed potential biases and assumption violations. Through careful validation and diagnostics, the BIC-selected model was identified as the final model, balancing simplicity with high predictive performance. This methodical process provides confidence in the model’s applicability and reliability for predicting **Quarterly_Average**.


# Results

#### 1. **Introduction to Results**
This section presents the comprehensive results of the multiple linear regression (MLR) analysis. It details the evaluation, refinement, and selection of the final model, emphasizing the predictive performance, key decisions, and diagnostic tests conducted to validate the model's robustness and reliability.


#### 2. **Model Comparison**
The analysis involved testing various models to balance predictive power and interpretability. Key model selection methods included:
- **AIC-based Stepwise Regression**: Focused on optimizing model fit while allowing for complexity.
- **BIC-based Stepwise Regression**: Prioritized simpler models to reduce overfitting risks.

```{r echo=FALSE}
# Load necessary library
library(knitr)

# Create a data frame with key metrics
comparison_table <- data.frame(
  Metric = c(
    "Adjusted R^2", 
    "Residual Standard Error (RSE)", 
    "RMSE (Cross-Validation)", 
    "MAE (Cross-Validation)", 
    "Predictors Selected"
  ),
  `AIC-Selected Model` = c(
    0.9937, 
    0.01029, 
    0.01088, 
    0.00918, 
    10
  ),
  `BIC-Selected Model` = c(
    0.9937, 
    0.01029, 
    0.01110, 
    0.00953, 
    10
  )
)

# Display the table using kable
kable(
  comparison_table, 
  col.names = c("Metric", "AIC-Selected Model", "BIC-Selected Model"),
  caption = "Comparison of AIC- and BIC-Selected Models",
  align = "lcc"
)

```

Both models performed similarly, with the AIC-selected model slightly outperforming on cross-validation metrics. 


#### 3. **Key Decisions**
To build a strong and reliable model, important features like `Detached_Unabsorbed_Quarterly_Avg` and `CPI_Quarterly_Avg` were kept because they were both statistically significant and theoretically important. Less useful features, like `Completed_Construction_Detached`, were removed to avoid overfitting the model. Transformations were used to improve accuracy: log transformations helped manage non-linearity and stabilize data for variables like `Detached_Unabsorbed_Quarterly_Avg`, while a Box-Cox transformation adjusted the response variable `Quarterly_Average` for better fit. For model selection, AIC was used to focus on minimizing errors, and BIC helped ensure the model remained simple and avoided overfitting.


#### 4. **Model Diagnostics and Assumptions**
To ensure the model's accuracy and reliability, several checks were performed. Residual analysis showed that the errors were mostly normal, as confirmed by the Q-Q plot, and there were no noticeable patterns in residual variance, confirming consistent variance (homoscedasticity). Multicollinearity was flagged between predictors like GDP_Quarterly_Avg and CPI_Quarterly_Avg using Variance Inflation Factors (VIFs), but these variables were kept because of their importance in the model. Influential data points, identified using Cook’s Distance, were removed to avoid skewed results. Overall, the model satisfied all key assumptions of linear regression, including linearity, normality, independence, and homoscedasticity, ensuring it is both valid and robust.


#### 5. **Interpretation of Final Model**
The final BIC-selected model included 10 significant predictors. Key metrics:  
- **Multiple \( R^2 \)**: 0.9946 (99.46% of the variability explained).  
- **Adjusted \( R^2 \)**: 0.9937 (robust against overfitting).  
- **Residual Standard Error**: 0.01029.  
- **Cross-Validation Results**:  
  - **RMSE**: 0.01110  
  - **MAE**: 0.00953  
  - **\( R^2 \)**: 0.9933  

**Significant Predictors**:  
- **Detached_Absorption_Quarterly_Avg** (\( \beta = -1.552 \times 10^{-5}, p < 0.001 \)): Negative impact.  
- **Detached_Unabsorbed_Quarterly_Avg** (\( \beta = 2.236 \times 10^{-5}, p < 0.001 \)): Positive impact.  
- **Semi_Unabsorbed_Quarterly_Avg** (\( \beta = -2.864 \times 10^{-5}, p < 0.05 \)): Negative impact.  
- **GDP_Quarterly_Avg** (\( \beta = -3.004 \times 10^{-7}, p < 0.001 \)): Negative impact.  
- **CPI_Quarterly_Avg** (\( \beta = 0.01315, p < 0.001 \)): Strong positive effect.  
- **Starting_Detached_Construction** (\( \beta = -4.393 \times 10^{-6}, p < 0.001 \)): Negative impact.  
- **Under_Construction_Detached** (\( \beta = 6.240 \times 10^{-6}, p < 0.001 \)): Positive impact.  
- **Under_Construction_Semi** (\( \beta = 2.180 \times 10^{-5}, p < 0.001 \)): Positive impact.  
- **Completed_Construction_Semi** (\( \beta = -1.250 \times 10^{-5}, p < 0.01 \)): Negative impact.  

#### 6. **Visual Representation**
- **Residuals vs. Fitted Values**: No patterns detected, confirming linearity.
- **Q-Q Plot**: Residuals followed a normal distribution.
- **Predicted vs. Actual**: Strong alignment along the diagonal, indicating high predictive accuracy.

#### 7. **Conclusion of Results**
The analysis focused on understanding the factors influencing new home prices, and evaluating a robust model to predict them. Using multiple linear regression, the final model highlighted several key predictors with significant effects on new home prices. For instance, unabsorbed homes had a strong positive impact. Conversely, absorption and completed construction negatively affected price. Macroeconomic indicators like GDP showed a negative effect, while CPI had a substantial positive impact. Construction activity variables, such as starting construction and under construction, further revealed nuanced effects on price trends. The model, explaining over 99% of the variability in new home prices, provides a comprehensive view of how inventory, construction activity, and macroeconomic factors collectively shape the housing market, offering valuable insights for policymakers and market stakeholders. Our model failed validation with our test and train datasets, indicating that it struggles to generalize accurately across different data subsets. This highlights potential limitations in its ability to make reliable predictions. However, it is important to note that the primary objective of this model is not solely predictive accuracy, but rather to provide insights into underlying behaviors and relationships within the data. Despite its shortcomings in predictive performance, the model's ability to identify significant patterns and key drivers of variability remains valuable for understanding the dynamics at play. 
# Conclusion and Limitations

```{r include-conclusion, echo=FALSE, eval=FALSE}
# Summarize:
# - How the final model answers the research question.
# - Broader significance of the model.
# Discuss:
# - Any unresolved limitations, such as violated assumptions or data issues.
```


# Ethics Discussion

```{r include-ethics, echo=FALSE, eval=FALSE}
# Discuss the choice of selection methods (manual or automated).
# Use material from the ethics module to defend your argument.
```

# References

```{r references, echo=FALSE, eval=FALSE}

```

