---
title: "STA302 Fall 2024 Final Project Part 1: Research Proposal and Data Introduction"
author: 
- "Wendy Huang"
- "Gadiel David Flores"

date: "07/10/24"
subtitle: "Research Proposal and Data Introduction"
output: pdf_document
format: pdf
number-sections: true
bibliography: references.bib
---
# Introduction
Housing affordability and price fluctuations have become pressing issues in metropolitan areas, and Toronto is no exception. As one of Canada's largest and fastest-growing cities, understanding the factors influencing house prices in Toronto is crucial for policymakers, developers, and potential homeowners. Over the past decade, housing prices have seen a steep upward trend, driven by a combination of housing supply and demand, economic, and policy-related factors.(@5causesbehindtoronto) This study aims to investigate the relationship between several key housing market indicators—namely, the number of available homes, homes sold, homes starting construction, new homes completed, and interest rate hikes—and housing prices in Toronto. By analyzing these factors, this research will contribute to a more comprehensive understanding of Toronto's housing market dynamics.

The rapid rise in housing prices has been widely debated, with supply constraints being one of the leading factors. For example, Glaeser, E. L., Gyourko, J., & Saks, R. E. (@whyhavehousingprices) employed linear regression to analyze the relationship between supply constraints and prices, showing an imbalance between supply and demand results in price hikes. This paper as well supports the use of housing starts and completions as critical indicators of price trends, aligning with this proposed research focus on construction-related factors in Toronto's housing market. 

Additional studies support the notion that housing availability and construction trends influence prices. In their paper, Goodman, A. C., & Thibodeau, T. G. (@metropolitanhousing) explored the relationship between new housing construction and price stability in urban areas. The authors found that increased housing completions have a stabilizing effect on prices, especially in high-demand areas. Their methodology relied on linear regression, confirming its suitability for modeling how newly completed homes impact price trends. 

The paper by Guiwen Bai and Yu Ma (@fundamentalhousing) uses a multi-factor regression model to study how these fundamentals, particularly the cost of borrowing and income tax impacts and interest rates, relate to housing price movements in Canada. The paper confirmed that low interest rates contribute to increased affordability, encouraging more people to enter the housing market.

Given these findings, linear regression is an appropriate tool for analyzing how these supply-side continuous variables interact to affect prices in Toronto. The clear quantitative relationship between supply predictors (available homes, homes sold, etc.) and price response aligns dynamically well with the assumptions of linear regression, making it a suitable method for this analysis. Linear regression allows us to quantify the relationship between these factors and price, providing insights into which variables have the most significant impact.

# Data Description

## Data Source

Using data from Statistics Canada, we extracted five datasets. @StatCan-NewHousingPriceIndex served as the foundation for this research paper, as it contained the main statistic we wanted to analyze. @StatCan-NewHousingPriceIndex represents the new housing price index, which calculates the average cost of new housing each month by multiplying by 100 and dividing by the base year. This provides an index reflecting housing costs for any given year and month. We then utilized @StatCan-CPI, @StatCan-GDP, @StatCan-HousingStarts, @StatCan-FinancialMarketStats, and @StatCan-HousingAbsorptions to extract our 13 predictors, including Absorption, GDP, CPI, interest rates, and Construction. Specifically, Absorption and Construction are further divided into two parts. Absorbtion is divided into semi-detached homes and single homes as well as empty and sold houses for any given month. Construction is categorized into three parts: new construction, under construction, and finished construction. These datasets were cleaned by removing missing values, converting data into whole integers, and merging them into one dataset. Additionally, the data spans from 1997 to 2016 and is divided quarterly, providing 76 entries. Lastly, we used interest rates to create our categorical predictor. This was done by comparing the entry i with entry i+1 in order to determine if there was an interest hike.   

## Response and Predictor Variables
Given our research question how do supply-side factors, such as housing availability, sales volume, construction starts, completions, and interest rate, impact housing prices in Toronto, the response variable chosen to help answer our research question is 'new housing price index'. This variable is suitable for linear regression because it is continuous, enabling us to model its relationships with multiple predictors. Additionally, since the new housing price index is influenced by economic variables, it allows us to identify trends and patterns through our linear regression model. For these reasons, it is a strong candidate for our model. Our five predictor variables are the number of available homes, homes sold, homes starting construction, new homes completed, and rate hikes

# Ethics Discussion
The datasets extracted from Statistics Canada are both trustworthy and ethically collected, as they originate from a reputable national statistical agency committed to maintaining data integrity and transparency. Statistics Canada follows strict protocols for data collection, ensuring that the information is accurate, reliable, and reflective of real market conditions (@StatCan). While the datasets themselves are typically reliable and collected following rigorous ethical standards, potential risks related to racial and ethnic disparities in housing must be acknowledged.

Ethical considerations also arise from the treatment of sensitive information, particularly regarding interest rates and housing data, which could impact individual financial circumstances. The datasets used in this study focus on aggregate data rather than individual responses, thus minimizing privacy concerns. 

Moreover, the datasets underwent thorough cleaning processes to ensure accuracy, further reinforcing their reliability for analysis. Overall, the ethical collection practices and transparency associated with these datasets support their trustworthiness for examining the impacts of supply-side factors on housing prices in Toronto.

# Preliminary Results 
Our model examines the relationship between the New Housing Price Index (NHPI) and predictors like detached absorption, unabsorbed homes, new homes under construction, completed constructions, and rate hikes. The intercept is 88.52, representing the NHPI value when all predictors are zero, serving as a baseline for evaluating the effects of other variables.

Detached absorption, our first predictor, shows that each additional home sold in a quarter is associated with a decrease of 0.0227 points in the NHPI. Conversely, unabsorbed homes increase the NHPI by 0.005315 points. The impact of new homes under construction is minimal, increasing the NHPI by 0.0009611 points for each home started. Completed constructions have a greater effect, contributing a 0.002977 point increase. Rate hikes, however, have the most significant influence: each rate increase leads to a 13.41-point decrease in the NHPI.

## Residual Analysis
We assessed the assumptions of our linear regression model by examining residual plots. The Residuals vs. Fitted Values plot (@fig-resid) reveals clustering around a fitted value of 100, with a slight decrease in spread, indicating potential non-linearity and correlated errors.

Further, we reviewed residuals vs. predictors. The residual plots for homes under construction and completed homes (@fig-residuals-3 and @fig-residuals-4) show minimal violations, as the spread appears random and well-distributed. However, the residuals for sold (@fig-residuals-1) and unsold homes (@fig-residuals-2) display fanning and clustering, suggesting correlated errors and non-constant variance. Additionally, the categorical predictor "rate hike" (@fig-residuals-5) shows wider residual variance for a value of "1" than for "0," indicating non-constant variance.

Lastly, while the QQ plot (@fig-qq_plot) suggests that most residuals align with the expected normal distribution, deviations at the lower and upper extremes indicate some skewness. This suggests that the residuals are not perfectly normally distributed, hinting at a potential violation of the normality assumption.

## Model Fitting {#sec-model}  

**Data**  
- **Quarterly Average**: This is our response variable, representing the average cost of a home in Canada.  
- **Detached Absorption Quarterly Avg**: This is our first predictor variable, describing the number of houses sold in a quarter.  
- **Starting Detached Construction**:  This predictor variable indicates the number of houses that began construction in a quarter.  
- **Completed Construction Detached**:  This represents the number of homes completed in a given quarter.  
- **Rates**:  This is our categorical predictor, with a value of 1 if there was a rate hike in that quarter or 0 if there was not.  
- **Detached Unabsorbed Quarterly Avg**:  This is our final predictor, indicating the number of available houses that were not sold during that quarter.  

```{r}
#| label: fig-data
#| fig-cap: "Data"
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readr)

# Load the dataset into R
data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

```{r}
#| label: fig-model
#| fig-cap: "Multiple Linear Model"
#| echo: false
#| warning: false
#| message: false

library(tibble)
library(knitr)

# multiple linear model
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg + 
              Detached_Unabsorbed_Quarterly_Avg + 
              Starting_Detached_Construction + 
              Completed_Construction_Detached + 
              rates, 
            data = data)

# create a data frame
coefficients_table <- as.data.frame(coef(model))
coefficients_table <- coefficients_table %>%
  rownames_to_column("Variable") %>%
  rename("Coefficient" = "coef(model)")

coefficients_table <- coefficients_table[1:6, ]

# summary table
kable(coefficients_table, caption = "Summary Table of Regression Coefficients", format = "markdown")
```

```{r}
#| label: fig-resid
#| fig-cap: "Residual vs Fitted Values"
#| echo: false
#| warning: false
#| message: false

# Residual and fitted values for the model
y_value <- resid(model)
x_value <- fitted(model)

# Plot the residual vs fitted values
plot(x = x_value, y = y_value, 
     main = "Residual vs Fitted", 
     xlab = "Fitted", 
     ylab = "Residuals")
```

```{r}
#| label: fig-residuals
#| fig-cap: "Residuals vs Each Predictor"
#| echo: false
#| warning: false
#| message: false

# Set up plot layout: 1 row, 3 columns
par(mfrow = c(1, 1))

# Residual vs Sold Homes
plot(x = data$Detached_Absorption_Quarterly_Avg, y = y_value, 
     main = "Residual vs Sold Homes", 
     xlab = "Sold Homes", ylab = "Residuals")

# Residual vs Unsold Homes
plot(x = data$Detached_Unabsorbed_Quarterly_Avg, y = y_value, 
     main = "Residual vs Unsold Homes", 
     xlab = "Unsold Homes", ylab = "Residuals")

# Residual vs New Homes Being Built
plot(x = data$Starting_Detached_Construction, y = y_value, 
     main = "Residual vs New Homes Being Built", 
     xlab = "New Homes Being Built", ylab = "Residuals")

# Residual vs Homes Completed
plot(x = data$Completed_Construction_Detached, y = y_value, 
     main = "Residual vs Homes Completed", 
     xlab = "Homes Completed", ylab = "Residuals")

# Boxplot of Residuals vs Rate Hike
boxplot(y_value ~ data$rates, 
        main = "Residual vs Rate Hike", 
        xlab = "Rate Hike", ylab = "Residuals")
```

```{r}
#| label: fig-qq_plot
#| fig-cap: "QQ Plot"
#| echo: false
#| warning: false
#| message: false

# Create a QQ plot for residuals
qqnorm(resid(model), main = "Normal QQ Plot of Residuals")
qqline(resid(model))
```
  
\newpage  

# Bibliography  