### Load Data
\begin{itemize}
  \item Quarterly_Average: the body mass index of an individual as the response
  \item Detached_Absorption_Quarterly_Avg: age of the individual (rounded to an integer) 
  \item Starting_Detached_Construction: Male or Female designation for gender (this survey predates when other options for this category were provided)
  \item Completed_Construction_Detached: a self-identified race with options White, Black, Hispanic, Mexican, Other
  \item rates: The midpoint of a household income bracket
  \item Detached_Unabsorbed_Quarterly_Avg: Hours of sleep per night on weekdays, self-reported
\end{itemize}
```{r}
# loads the dataset into R
data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

```
### Multiple Linear Model
```{r}
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg + 
              Detached_Unabsorbed_Quarterly_Avg + 
              Starting_Detached_Construction + 
              Completed_Construction_Detached + 
              rates, 
            data = data)
intercept <- round(as.numeric(coef(model)[1]))
```

### Assumptions of this model

```{r}
# replace NULL with appropriate values to be used in y and x axes of plot respectively.
y_value <- resid(model)
x_value <- fitted(model)

# plots the residual vs fitted plot
plot(x = x_value, y = y_value, main="Residual vs Fitted", xlab="Fitted",
     ylab="Residuals")
```
## Residuals vs each predictor
```{r}
# Plot multiple graphs in a single grid (1 row, 3 columns)
par(mfrow=c(1,1))

# Plot Residual vs Sold Homes
plot(x = data$Detached_Absorption_Quarterly_Avg, y = y_value, main="Residual vs Sold Homes", xlab="Sold Homes", ylab="Residual")

# Plot Residual vs Unsold Homes
plot(x = data$Detached_Unabsorbed_Quarterly_Avg, y = y_value, main="Residual vsUnsold Homes", 
     xlab="Unsold Homes", ylab="Residual")

# Plot Residual vs New homes being built
plot(x = data$Starting_Detached_Construction, y = y_value, main="Residual vs New homes being built", 
     xlab="New homes being built", ylab="Residual")

# Plot Residual vs Homes Completed
plot(x = data$Completed_Construction_Detached, y = y_value, main="Residual vs Homes Completed", 
     xlab="Homes Completed", ylab="Residual")

# Boxplot Residual vs Rate Hike
boxplot(y_value ~ data$rates, main="Residual vs Rate Hike", 
        xlab="Rate Hike", ylab="Residuals")
```
### QQ Plot
```{r}
# Create QQ plot for residuals
qqnorm(resid(model), main = "Normal QQ Plot of Residuals")
qqline(resid(model))
```

